---
title: "Quarto template for the _Journal of Fish and Wildlife Management_"
thanks: This draft manuscript is distributed solely for purposes of scientific peer review. Its content is deliberative and predecisional, so it must not be disclosed or released by reviewers. Because the manuscript has not yet been approved for publication by the U.S. Geological Survey (USGS), it does not represent any official USGS finding or policy.
author: "**Richard A. Erickson,* Althea A. Archer, Michael N. Fienen**"
date: today
format:
  docx: 
    fig-cap-location: bottom
    keep-tex: true
    reference-doc: jfwm_template.docx

csl: council-of-science-editors-author-date.csl 
bibliography: references.bib
header-includes:
  \usepackage{setspace}\doublespacing
  \usepackage{lineno} \linenumbers
  \usepackage{hanging}

editor_options: 
  markdown: 
    wrap: sentence
nocite: |
  @baumer2015r, @Leisch_2002, @rstudio
---

<!-- Note the use of two spaces to form a list without indents, from  -->
<!-- https://stackoverflow.com/questions/32474227/lists-without-bullets-points-in-markdown -->

**R.A. Erickson**  
U.S. Geological Survey, Upper Midwest Environmental Sciences Center, La Crosse, WI  
**A.A. Archer**  
U.S. Geological Survey, Water Resources Mission Area, Reston, VA  
**M.N. Fienen**  
U.S. Geological Survey, Upper Midwest Water Science Center, Madison, WI  
*Corresponding author: (rerickson@usgs.gov)  


# Abstract

Scientists and resource managers increasingly use Markdown-based tools to create reports and manuscripts.
We present a Quarto template for this journal, _Journal of Fish and Wildlife Management_, in our article.
We also provide a high-level overview of Quarto and other Markdown-based workflows as part of the document.
Lastly, we provide examples of some features of the Quarto publishing system that may be helpful for authors in this journal.  
**Keywords:** Findable, Accessible, Interoperable, and Reuseable (FAIR); markdown; reproducible science; journal template; Quarto

# Introduction 

Modern computing tools allow scientists and practitioners to be efficient, transparent, and reproducible [@Braga_2023; @Erickson_2021].
For example, ecologists are moving away from point-and-click statistical methods and using scripting languages with code [@borregaard2016towards].
These changes not only make scientists more productive but also confer additional benefits.
First, scripting allows methods to be reused by their creator.
For example, scientists commonly complete the same or similar analyses on a regular basis [such as @usfws2023], and scripting allows methods to be easily adapted and often quickly rerun with minimal formatting.
Second, sharing the scripts allows others to reuse methods and recreate results when data are shared.
Third, scripting documents exactly what options were, or were not, selected for an analysis and data cleaning.
All three of these features increase transparency in the scientific process.

The broader scientific community increasingly has begun to see the importance of reproducible results because the perception of a "reproducibility crisis" exists among most of the scientists surveyed across scientific fields [@Baker_2016; @fanelli2018science]. 
This lack of reproducibility also exists in natural resource fields. 
For example, @ArchMiller_2020 determined only 17% of studies randomly selected from the _Journal of Wildlife Management_ and the _Wildlife Society Bulletin_ could have their analyses reproduced.
In hydrology, lack of reproducibility has raised questions of the validity of the field as a whole [@Hutton_2016] and has motivated exploration of challenges from other fields [@Fienen_2016].
Thus, scientists increasingly seek to produce reproducible results, especially for computationally intensive projects [@borregaard2016towards].

Others have provided suggestions and methods for creating reproducible results in natural resource and related fields [such as @borregaard2016towards; @Braga_2023].
Additionally, Findable, Accessible, Interoperable, and Reusable (FAIR) principles exist to help scientists share and produce reproducible science [for example, @Reiser_2018].
Likewise, professional societies and government agencies have produced guidelines similar to FAIR for reproducible science (for example, see Table 1 in @Erickson_2021 that lists example organizations).

Many scientists use scripting languages like Python or R for their statistical methods and analyses. For example, @Erickson_2020 determined that a plurality of authors in the journal _Environmental Toxicology and Chemistry_ from 2019 used the R computing language for analysis in their papers.
However, we have observed that fewer scientists use similar tools for writing manuscripts even though scripting environments are now enriched with code-based manuscript services, such as Quarto, knitr, and Bookdown, which are primarily Markdown-based programs (Baumer and Udwin 2015; Figure 1).
Advantages of Markdown-based programs over direct typesetting products like LaTeX, Word, or Pages include simplicity, readability, transparency, and embedded computation [@baumer2015r]. Specifically, code-based manuscript writing can include the option to readily incorporate formatting from templates, ease of changing templates and reference formatting, automatic populating of references based upon Digital Object Identifiers (DOIs), and the ability to easily handle complex formulas (specifically embedded LaTeX syntax for formulas). 
Furthermore, these writing tools can be especially powerful when embedded with the code and data used in the document, which allows authors to create reproducible manuscripts that directly populate in-text tables and figures. 

On a more personal note and of possible interest to readers of this journal, we have observed interest (circa 2021-2023) in Markdown-based writing documents from our colleagues and collaborators who are natural resource managers or scientists supporting these managers, in North America and globally.
Furthermore, these examples motivated our creation of this paper and corresponding template. 
For example, a U.S. Geological Survey Markdown Community of Practice exists and all of the authors have presented tutorials to this group.
Additionally, this group includes members from the U.S. Fish and Wildlife Service and other agencies in the U.S. Department of the Interior.
We have also observed conversations about using Markdown-based writing tools in online conversations such as the U.S. Fish and Wildlife Service's "Great Lakes R Working Group."

For readers who are unfamiliar with Markdown-based writings tools, we provide a brief history here.
Stanford University computer scientist Donald Knuth had trouble typesetting his mathematical equations in the late 1970s and early 1980s and this led him to create the TeX language [@knuth1984tex].
Building upon TeX, Leslie Lamport created LaTeX as a set of macros to make TeX easier to use [@lamport1985latex].
Statisticians who used LaTeX sought to readily embed statistical code, results, and figures with LaTeX documents.
Thus, to meet this need, the program Sweave was created to "weave" S into LaTeX (at the time, many statisticians used the S language, which inspired the R language; Leisch 2002).

Although brilliant in theory, we have observed that TeX, LaTeX, and Sweave all can be complicated to use, and Sweave can be especially cumbersome because it requires the source file to be compiled multiple times for a single output. 
TeX, LaTeX, and Sweave also may be difficult for natural resource scientists to learn, because they require new software and languages outside of the more common R or Python environments.
Furthermore, in the four and half decades since Tex was created and four decades since LaTeX was created, Markdown usage became ubiquitous through its integration with version control tools such as Git and common Microsoft Office tools such as Teams and Word, which include many common Markdown shortcuts for formatting text (for example, typing `*` then `space` on a new line in either program will start a bulleted list, which is syntax from the Markdown language).
Others noted these difficulties as well; thus Yihui Xie and others created a Markdown-based program for use with R, known as R Markdown [@Xie_2018], that overcomes many of these limitations.
R Markdown works by using the `knitr` package to "knit" (or compile) the R Markdown into an output such as a Hyptertext Markup Language files, ending with `.html`; Microsoft Word Document, ending with `.docx`; or Portable Document Format, ending with `.pdf` [@knitr2014].
Another Markdown offshoot is the ROxygen2 language, which allows users in the R environment to directly convert R script files (ending with `.R` rather than R Markdown files ending with `.Rmd`) into R Markdown, which is then "knit" to output formats mentioned in the previous sentence [@roxygen2].
The broader R community received R Markdown well [such as reviews by @baumer2015r], but this language also has some limitations.
Most notably, the language requires R and has some minor quirks.
The popularity and limitations of R Markdown led a group including Yihui Xie to create a more general program, Quarto, that works natively with many languages rather than only R [@Allaire2022].

The remaining purpose of our paper is to present our Quarto template for this journal, which can be fully accessed at @erickson2023quarto.
We demonstrate some useful features including how to include figures, tables, and the `quarto-utils` package for automatic bibliography generation.
We also discuss how these tools may be applied to other situations.
Although our examples tend to focus on R, Quarto works with many languages including LaTeX (for example, to typeset equations and format outputs), Python, and Julia (Figure 1).

# Methods

## Quarto workflow

This tutorial focuses on Word document outputs because the _Journal of Fish and Wildlife Management_ uses a Word-based workflow.
When writing a Quarto document, the first step is to create a Quarto file that ends with `.qmd` extensions.
Common editors that work with Quarto files include Posit RStudio (v2023.06 or newer; Posit Team 2023) and Microsoft Visual Studio Code [@code2023visual].
RStudio also offers the benefits of creating new Quarto files from templates for journal articles, presentations, and other types of documents directly within the editor.
Additionally, Quarto contains tools for generating empty templates outside of RStudio using command line tools for editing in programs such as Visual Studio Code or similar editor programs (see the program documentation such as https://quarto.org/docs/journals/templates.html for addition details of these features, accessed 8 December 2023).
Alternatively, readers may use an existing Quarto file and edit that file.
For example, people can edit the Quarto file used to generate this paper that is hosted at @erickson2023quarto to write their own _Journal of Fish and Wildlife Management_ manuscripts.

Within the Quarto file, the header contains metadata in a YAML-style format.
The Quarto file for this article contains the following metadata:

* The `title` listing the article's title.
* The `author` listing the article's author or authors.
* The `date` the Quarto file was last rendered to create the Word file.
* The output `format` options including the output type of `docx`, an option to set the figure caption locations (`fig-cap-location`) to the `bottom`, an option to keep the LaTeX output files (`keep-tex: true`), and the Word file used as the style template for the file (`reference-doc: jfwm_template_template.docx`).
* The `csl` file listing article's citation style; for the _Journal of Fish and Wildlife Management_, this is the "Council of Science Editors, author-date format indents" (`council-of-science-editors-author-date.csl`). "Council of Science Editors, author-date format indents" refers to the "Council of Science Editors" formatting style using the Council of Science Editors style with authors and dates (such as "example fact _Author et al. 2023_" and in contrast to other formats numeric such as "fact [1]") and indented reference section paragraphs.
* The name and location of the reference file (`bibliography: references.bib`).
* Code to include LaTeX-based headers (`header-includes:`) for double spacing (`\usepackage{setspace}\doublespacing`) and line numbers (`\usepackage{lineno} \linenumbers`).
* Editor options (`editor_options`) telling the Markdown editor (`markdown`) to use a soft line wrap after each sentence (`wrap: sentence`).
* A list of "no citation" references (`nocite`) tells the Markdown to include citation items even if the item is not cited in the document. We include this because we had to manual format some inline citations to match the journal's style.

After the header, the user writes their manuscript using Markdown code with optional LaTeX code for equations. 
Code can be embedded within "chunks" that are defined by their language (such as `r` or `python`). These code chunks can either be displayed or hidden in the final document.
References are added to the BibTeX file, either manually or using a script described later by pulling DOIs.
Figures and tables can be added as needed, whether built locally with code chunks or called in from other sources.
While writing, the author can preview the results by compiling the file to create a Word document.
This also serves as a check for code mistakes that might prevent the Quarto file from compiling.
The end result is a Word file that may be submitted to the _Journal of Fish and Wildlife Management_ with minimal manual formatting required. 

## Background knowledge, tips, and other Quarto observations 

Writing Quarto documents requires a basic knowledge of Markdown.
The Quarto documentation (available at [https://quarto.org/](https://quarto.org/), accessed September 2023) contains tutorials to help people get started.
Additionally, the Quarto document [@erickson2023quarto] used to generate this document contains many examples for those who like to dive in and learn by doing.
Writing equations requires an understanding of some of the basics of LaTeX but much less than writing entire documents in LaTeX (many online tutorials exist such as the TeX User Group's, http://tug.ctan.org/info/short-math-guide/short-math-guide.pdf, accessed 8 December 2023).


As a general observation, when Quarto documents are rendered to Word documents, the Pandoc software [@dominici2014overview] that renders the document is limited in formatting and style options compared to Hypertext Markup Language or Portable Document Format files because of technical reasons beyond the scope of this paper.
Hence, readers may find some solutions using search engines that will not work for this Word-based template.
For example, cross-references for figures and tables do not work with this template.
Likewise, we had to use `nocite` and manually type in some references due to formatting challenges with the use certian species characters such as semicolons.
Additionally, readers familiar with Quarto may find some quirky examples in the Quarto template to help meet the _Journal of Fish and Wildlife Management_ formatting style.
For example, the authors' names are all listed on one line rather than using multiple entries as one might expect in the metadata section.

```{r}
#| echo: false
n_total <- 4341
n_mean <- n_total / 23.5
```
Integrating code into Quarto documents also helps to create reproducible results.
Code may be included directly within the file, or R script files may be called into the Quarto file and "knit" into the final outputs using the `read_chunk()` function from the `knitr` package [@knitr2014].
For example, a simple regression might be run or data plotted and then the outputs included in the file.
Including the code ensures the manuscript always has the most recent results and statistics if the input files are changed.
Additionally, code outputs can be included and displayed inline with R (this is currently under development with Python, see [https://github.com/quarto-dev/quarto-cli/pull/6190](https://github.com/quarto-dev/quarto-cli/pull/6190), accessed September 2023).
For example, when describing a study or writing a monitoring report, one might write "We observed many birds (_n_ = ``r prettyNum(n_total, big.mark= ",")``) at the refuge with a mean  of ``r round(n_mean, 1)`` birds per day."
With this example, inline code would allow these numbers to updated dynamically if the source data or code changed when the file is rerendered.
Hence, if the authors updated their input numbers in either the source data files or code, the numbers in text would change.
This change may occur because the authors updated their data while writing the manuscript, or because the authors update their report on a regularly basis (such asa report an annual monitoring program). 

Another tip for using Quarto to render Word documents is to use the `flextable` package in R [@flextable2013] for creating tables rather than the popular `kable()` function from the `knitr` package [@knitr2014] and `kableExtra` package [@kableExtra] because the `flextable` package more readily allows tables to be formatted in Word-based outputs.
For example, we demonstrate how the `flextable` package may be used to create simple tables in Table 1 and tables with multiple columns and rows in Table 2. 
This table has multicolumn headers and automatically drops duplicate row names.
These two aesthetic features are commonly used by scientists when describing data and can be automatically created using the `flextable` package in R.
These features exist  with `kable`-type tables, but do not render correctly in Word document outputs.

Finally, to edit the text style, change the `jfwm_template.docx` file using Word rather than directly editing the output rendered from Quarto code.
This may require practitioners to become familiar with Word templates.
For example, in the _Journal of Fish and Wildlife Management_, the first paragraph of every section lacks indentation, but the second and subsequent paragraphs in a section use a tab indent. 
Changing this formatting requires changing the paragraph template settings in the template Word document.
Further modifications can be made to change the template by updating the Styles definitions (such as "Header 1" or "Body Text") in the template Word document.
These changes would then be shown on the output Quarto-based Word document after the template Word document is saved and  the Quarto (`.qmd`) fille is rerendered to create Word document.

## `quarto-utils`

The `quarto-utils` software is a Python package that assists Quarto authors in compiling and formatting references. 
By leveraging the https://doi.org web interface, authors are able to place references in their documents by only providing a DOI in place of a reference in the text.
Then using functionality in `quarto-utils`, the code will search the `.qmd` document for DOI codes, retrieve a BibTeX representation of the reference from doi.org, add the BibTeX string to a bibliography file, and convert the DOI reference in the `.qmd` file to a valid reference call.
The author identifies a reference in the same way they would call out an existing reference using `[]` delimiters, but rather than providing a key that references the `.bib` file (such as `[@Prince1999]`), they would use either `_doi:thedoinumber` or `\_doi:thedoinumber` in its place.
The `quarto-utils` code searches for `_doi:` or `\_doi:` and parses for the DOI.
The DOI can take the form of just the number (such as  `10.1999/prince.albums`) or the full uniform resource locator (such as `https://doi.org/10.1999/prince.albums`).
If invalid DOI codes are provided, `quarto-utils` will report that it was not able to find a reference for a given string and will leave the `.qmd` file unchanged.
This feature is another method to streamline the writing process.
Many bibliographic database software programs exist, but authors may find it more efficient to simply search the internet for a reference they want to cite, copy the DOI, paste it into their `.qmd` file, and keep writing.
The BibTeX file generated by `quarto-utils` may require manual formatting, and occasionally, automatically generated outputs prevent Quarto from rendering files until special characters (such as non-American Standard Code for Information Interchange) are removed.
Also, citations listed on doi.org may be incorrect and sometimes requires addition curation and formatting.

# Results

We have demonstrated how to use Quarto to prepare manuscripts for the _Journal of Fish and Wildlife Management_, including figures (such as Figure 1) and tables (such as Table 1 and Table 2).
We have also described the coauthors' serendipitous connections to the Upper Midwest (Table 2).
The methods of this paper provided tips and suggestions for using Quarto to format documents to match this journal's style. 
Last, we presented use a Python package, `quarto-utils` for helping author automate their reference management. 


# Discussion

Script-based writing programs such as Quarto offer advantages compared to traditional word processing programs by allowing writers to focus on writing rather than formatting [@baumer2015r].
For example, scientists can readily change formatting if they need to change reports (such as completion reports or partner summaries) to journal articles or change formatting across journal articles because only style files need to be changed rather than reformatting the entire document.
Additionally, reports that are completed and rerun on a regular basis can be formatted to automatically include numbers in the text to avoid tedious and error prone copying and pasting of values such as the examples previously shown in this document where numbers are reinserted each time the source file is recompiled.
When integrated with statistical programs such as Python or R and also publicly released, files created these programs also more readily support FAIR principles because the details (and code) necessary to recreate results are embedded within the Quarto document.
The FAIR principles are increasingly becoming used by science agencies for conducting and documenting research and monitoring programs.
For example, the U.S. Geological Survey includes FAIR principles as part of their road map for science [@lightsom2022opportunities] and U.S. Fish and Wildlife Service planning documents also includes FAIR principles [@usfws2023budget].

Thus, we created an example of how to prepare manuscripts in a Quarto document for the _Journal of Fish and Wildlife Management_.
Aside from directly helping potential authors in this journal, we also hope our publication will raise awareness for fish and wildlife managers about this tool.
For example, many reports that are published weekly, monthly, quarterly, annually, or on a similar interval would benefit from not only using a computer script to run the statistical analysis, data formatting, plotting, or other number crunching but also in the document preparation itself.
The example described in this paper and fully provided in @erickson2023quarto could also serve as a starting place for these documents.

# Conflict of Interest

The authors report no conflicts of interest.

# Acknowledgments

We thank the U.S. Geological Survey Biological Threats and Invasive Species Research Program and U.S. Geological Survey Water Resources Mission Area Integrated Information Dissemination Division for funding.
Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
The findings and opinions expressed in this manuscript are those of the authors and do not necessarily represent the views of the U.S. Fish and Wildlife Service.

# Data Availability

The template code for the Quarto document used to create this article is available in @erickson2023quarto.

# Figures

![Figure 1. Example of Quarto workflow. Within the Quarto document, users can add chunks of code in the R, Julia, or Python languages. In addition, users can directly embed LaTeX typesetting, which is often used for equations. Quarto then renders the document to an intermediate Markdown document with Pandoc. Pandocs then converts the Markdown document directly to one of the many output formats. These formats include such products as Portable Document Format (PDF), Microsoft Word, Microsoft PowerPoint, or Hypertext Markup Language (html).](quarto.png)


# Tables

Table 1. Steps to use the Quarto template associated with this manuscript for a _Journal of Fish and Wildlife Management_ article.  
```{r}
#| message: false
#| warning: false
#| error: false
#| echo: false
#| eval: true
library(tidyverse)
library(flextable)

## From https://www.reddit.com/r/rstats/comments/11ywud9/how_do_you_get_perfect_tables_for_microsoft_word/
# Function to have tables autofit to page
FitFlextableToPage <- function(ft, pgwidth = 6.5){
  # set as autofit to make width parameters adjustable
  ft_out <- ft %>% autofit()
  # set width as function of page width
  ft_out <- width(ft_out,
                  width = dim(ft_out)$widths * pgwidth /
                    (flextable_dim(ft_out)$widths))
  return(ft_out)
}

steps_tib <- tibble(
  order = c(1, 2, "3 (optional)", 4, "5 (optional)", 6),
  step = c("Obtain template from DOI",
           "Write text using Markdown",
           "Run Python script to populate bibliography file",
           "Render document to Word",
           "Break Quarto-Word link and do final formatting manually",
           "Submit to journal"))

steps_tib |>
  flextable() |>
  theme_booktabs() |>
  FitFlextableToPage()
```  
Table 2. Author trivia to demonstrate a table. The table awkwardly includes wide and long form data to demonstrate how to use the `flextable` package.  
```{r}
#| message: false
#| warning: false
#| error: false
#| echo: false
#| eval: true

data <-
  tibble(Author = c("RAE", "RAE", "RAE", "AAA", "MNF"),
         Internship = c("University of Minnesota", "--", "--",
                       "University of Minnesota", "--"),
         Undergraduate = c("--", "--", "--", "Gustavus Adolphus College", "Macalester College"),
         "University of Wisconsin (UW) System" =
           c("Undergraduate, UW-Stevens Point",
             "Internship, UW-Madison",
             "Graduate faculty, UW-La Crosse",
             "Graduate Certificate, UW-Madison",
             "Graduate faculty, UW-Madison"))

data |>
  flextable() |>
  theme_booktabs() |>
  add_header_row(
    colwidths = c(1, 2, 1),
    values = c("",
               "Minnesota Connection", "Wisconsin Academic Connection")) |>
  merge_v(j = ~Author) |>
  FitFlextableToPage()
```

# Supplemental Material

**Supplement S1**  U.S. Fish and Wildlife Service. 2023a. Waterfowl population status, 2023. U.S. Department of the Interior, Washington, D.C. USA.  
**Supplement S2** U.S. Fish and Wildlife Service. 2023b. Budget justifications and performance information fiscal year 2024. U.S. Department of the Interior, Washington, D.C. USA.

# References
